---
title: "Exploration and Modeling of Warfarin Dosage"
output: html_document
---
Ploypun Narindrarangkura
Bioinformatics
Data Analysis using R
December 7, 2017

A warfarin dosing estimator (http://warfarindosing.org/Source/Home.aspx) was developed to a guide for use by clinicians in beginning warfarin therapy for patients where medication with a blood thinner is advised to reduce the risk of stroke, heart attack, or other serious conditions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#1. Perform data cleaning.
```{r}
setwd("~/desktop")
mydata <- read.delim("~/Desktop/iwpc_warfarin.txt")

#rename header
colnames(mydata)[1] <- "SubjectID"
colnames(mydata)[2] <- "SampleID"
colnames(mydata)[3] <- "Project"
colnames(mydata)[4] <- "Gender"
colnames(mydata)[6] <- "Race"
colnames(mydata)[8] <- "Ethnicity"
colnames(mydata)[9] <- "Age"
colnames(mydata)[10] <- "Height"
colnames(mydata)[11] <- "Weight"
colnames(mydata)[12] <- "Indication"
colnames(mydata)[15] <- "Target_INR"
colnames(mydata)[16] <- "INR_Range"
colnames(mydata)[17] <- "Subject_reached_dose"
colnames(mydata)[18] <- "Dose_Warfarin"
colnames(mydata)[19] <- "INR"
colnames(mydata)[20] <- "Smoking"
colnames(mydata)[21] <- "CYP2C9"
colnames(mydata)[22] <- "VKORC1"

mydata$Medications <- as.character(mydata$Medications)
mydata$Comorbidities <- as.character(mydata$Comorbidities)

#Delete columns
mydata <- mydata[, -5]
mydata <- mydata[, -6]

#Change class
mydata$Project <- as.factor(mydata$Project)
mydata$Subject_reached_dose <- as.factor(mydata$Subject_reached_dose)
mydata$Smoking <- as.factor(mydata$Smoking)

#Age
summary(mydata$Age)
levels(mydata$Age)[1] <- "10 - 19"

#rename value in Age
levels(mydata$Age)[1] <- "1"
levels(mydata$Age)[2] <- "2"
levels(mydata$Age)[3] <- "3"
levels(mydata$Age)[4] <- "4"
levels(mydata$Age)[5] <- "5"
levels(mydata$Age)[6] <- "6"
levels(mydata$Age)[7] <- "7"
levels(mydata$Age)[8] <- "8"
levels(mydata$Age)[9] <- "9"
mydata$Age <- as.numeric(mydata$Age)

#rename value in Smoking
levels(mydata$Smoking)[1] <- "No"
levels(mydata$Smoking)[2] <- "Yes"

#Cleaning INR Range
summary(mydata$INR_Range)
levels(mydata$INR_Range)[levels(mydata$INR_Range)=="3-Feb"] <- "2-3"

mydata$INR_Range[mydata$INR_Range == levels(mydata$INR_Range)[7]] <- levels(mydata$INR_Range)[6]

#Drop unused factor levels: in this case will drop level 4-Mar
mydata$INR_Range <- droplevels(mydata$INR_Range)

#Ordering INR range
mydata$INR_Range <- factor(mydata$INR_Range, levels=c("1.7-2.8", "1.7-3.3", "2-3", "2-3.5", "2.5-3.5", "3.0-4.0"))

#Ordering VKORC1
mydata$VKORC1 <- factor(mydata$VKORC1, levels=c("G/G", "A/G", "A/A"))
```

Medication
```{r}
#Amiodarone/Cordarone/Pacerone
library(stringr)
library(plyr)
mydata$amiodarone <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "amiodarone"),
                            yes = 1,
                            no = 0)
mydata$amiodarone[str_detect(mydata$Medications, "not amiodarone")==T] = 0
mydata$amiodarone[str_detect(mydata$Medications, "no amiodarone")==T] = 0

#Aspirin
mydata$aspirin <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "aspirin"),
                             yes = 1,
                             no = 0)
mydata$aspirin[str_detect(mydata$Medications, "not aspirin")==T] = 0
mydata$aspirin[str_detect(mydata$Medications, "no aspirin")==T] = 0

#Statin
mydata$statin <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "simvastatin|Simvastatin|atorvastatin"),
                             yes = 1,
                             no = 0)
mydata$statin[str_detect(mydata$Medications, "not simvastatin")==T] = 0
mydata$statin[str_detect(mydata$Medications, "no simvastatin")==T] = 0

mydata$statin[str_detect(mydata$Medications, "not Simvastatin")==T] = 0
mydata$statin[str_detect(mydata$Medications, "no Simvastatin")==T] = 0


#Azoles
mydata$azoles <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "azole"),
                              yes = 1,
                              no = 0)
#Sulfa
mydata$sulfa <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "Septra|bactrim"),
                         yes = 1,
                         no = 0)

#Acetaminophen
mydata$acetaminophen <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "acetaminophen|Acetaminophen"),
                              yes = 1,
                              no = 0)

mydata$acetaminophen[str_detect(mydata$Medications, "not acetaminophen")==T] = 0
mydata$acetaminophen[str_detect(mydata$Medications, "no acetaminophen")==T] = 0

#Clopidogrel
mydata$clopidogrel <- ifelse( !is.na(mydata$Medications) & str_detect(mydata$Medications, "clopidogrel"),
                          yes = 1,
                          no = 0)

#Enzyme Inducers
mydata$enzyme_inducer <- ifelse(!is.na(mydata$Medications) & str_detect(mydata$Medications, "carbamazepine|Carbamazepine|phenobarbital|phenytoin|Phenytoin|rifampin"),
                              yes = 1,
                              no = 0)

mydata[,21:28] <- lapply(mydata[,21:28], as.factor)
```

Impute missing data
```{r}
mydata$Weight[is.na(mydata$Weight)] <- mean(mydata$Weight,na.rm=T)
mydata$Height[is.na(mydata$Height)] <- mean(mydata$Height,na.rm=T)
mydata$INR[is.na(mydata$INR)] <- mean(mydata$INR,na.rm=T)

mydata <- mydata[, -c(1:3)]
mydata <- mydata[, -c(7:9)]
mydata <- mydata[, -c(7:9)]
mydata <- mydata[, -9]

summary(mydata)
```

#2.Create two subsets for further exploration.
```{r}
#Warfarin level
cuthigh <- mean(mydata$Dose_Warfarin, na.rm = TRUE) + 
  0.2*sd(mydata$Dose_Warfarin, na.rm = TRUE)

cutlow <- mean(mydata$Dose_Warfarin, na.rm = TRUE) - 
  0.2*sd(mydata$Dose_Warfarin, na.rm = TRUE)

mydata$warfarin_level <- mydata$Dose_Warfarin

mydata$warfarin_level <- cut(mydata$warfarin_level, breaks=c(0, cutlow, cuthigh, Inf))
levels(mydata$warfarin_level)[1] <- "Low dose"
levels(mydata$warfarin_level)[2] <- "Medium dose"
levels(mydata$warfarin_level)[3] <- "High dose"
```

Subset data
```{r}
HighDose <- subset(mydata, warfarin_level == "High dose")
MediumDose <- subset(mydata, warfarin_level == "Medium dose")
LowDose <- subset(mydata, warfarin_level == "Low dose")
```

#3. Display aggregate information on the data.
```{r}
summary(HighDose)
summary(LowDose)
```

#4. Create visualizations of the distributions of key variables by high/low warfarin dosage.
```{r}
library(ggplot2)
library(Rmisc)
```
GENDER
```{r}
g1 <- ggplot(data=subset(HighDose, !is.na(Gender)), main = "High Dose Warfarin")
g2 <- ggplot(data=subset(LowDose, !is.na(Gender)), main = "Low Dose Warfarin")
p1 <- g1 + geom_bar(aes(Gender, fill=Gender)) +
  xlab("Gender") +
  ylab("Frequency") +
  ggtitle("High Dose Warfarin")
p2 <- g2 + geom_bar(aes(Gender, fill=Gender)) +
  xlab("Gender") +
  ylab("Frequency") +
  ggtitle("Low Dose Warfarin") 
multiplot(p1, p2, cols=2)
```

RACE
```{r}
g3 <- ggplot(data=subset(HighDose, !is.na(Race)), main = "High Dose Warfarin")
g4 <- ggplot(data=subset(LowDose, !is.na(Race)), main = "Low Dose Warfarin")
p3 <- g3 + geom_bar(aes(Race, fill=Race)) +
  xlab("Race") +
  ylab("Frequency") +
  ggtitle("High Dose Warfarin")
p4 <- g4 + geom_bar(aes(Race, fill=Race)) +
  xlab("Race") +
  ylab("Frequency") +
  ggtitle("Low Dose Warfarin") 
multiplot(p3, p4, cols=2)
```

ETHNICITY
```{r}
g5 <- ggplot(data=subset(HighDose, !is.na(Ethnicity)), main = "High Dose Warfarin")
g6 <- ggplot(data=subset(LowDose, !is.na(Ethnicity)), main = "Low Dose Warfarin")
p5 <- g5 + geom_bar(aes(Ethnicity, fill=Ethnicity)) +
  xlab("Ethnicity") +
  ylab("Frequency") +
  ggtitle("High Dose Warfarin")
p6 <- g6 + geom_bar(aes(Ethnicity, fill=Ethnicity)) +
  xlab("Ethnicity") +
  ylab("Frequency") +
  ggtitle("Low Dose Warfarin") 
multiplot(p5, p6, cols=2)
```

AGE
```{r}
g7 <- ggplot(data=subset(HighDose, !is.na(Age)), main = "High Dose Warfarin")
g8 <- ggplot(data=subset(LowDose, !is.na(Age)), main = "Low Dose Warfarin")
p7 <- g7 + geom_bar(aes(Age, fill=Age)) +
  xlab("Age in Decades") +
  ylab("Frequency") +
  ggtitle("High Dose Warfarin")
p8 <- g8 + geom_bar(aes(Age, fill=Age)) +
  xlab("Age in Decades") +
  ylab("Frequency") +
  ggtitle("Low Dose Warfarin") 
multiplot(p7, p8, cols=2)
```

#5. Create visualizations of a couple of relationships you find interesting between variables.

HEIGHT AND DOSE
```{r}
g9 <- ggplot(data=mydata)
g9 + geom_point(data=subset(mydata, !is.na(Height) & !is.na(Dose_Warfarin)), aes(y=Height, x=Dose_Warfarin, color=warfarin_level), size=0.5) +
  xlab("Warfarin Dosage") +
  ylab("Height (cm)") +
  ggtitle("Scatter plot")

g9 + geom_boxplot(data=subset(mydata, !is.na(Height) & !is.na(warfarin_level) ), aes(y=Height, x=warfarin_level, fill=warfarin_level), alpha =.5) +
  xlab("Warfarin level") +
  ylab("Height (cm)") +
  ggtitle("Box plot")
```

WEIGHT AND DOSE
```{r}
g9 + geom_point(data=subset(mydata, !is.na(Weight) & !is.na(Dose_Warfarin)), aes(y=Weight, x=Dose_Warfarin, color=warfarin_level), size=0.5) +
  xlab("Warfarin Dosage") +
  ylab("Weight (kg)") +
  ggtitle("Scatter plot")

g9 + geom_boxplot(data=subset(mydata, !is.na(Weight) & !is.na(warfarin_level) ), aes(y=Weight, x=warfarin_level, fill=warfarin_level), alpha =.5) +
  xlab("Warfarin level") +
  ylab("Weight (kg)") +
  ggtitle("Box plot")
```

#6. Fit 2 regression models, as described below, to estimate warfarin dosage.
```{r}
mydata <- na.omit(mydata)

model1 <- lm(Dose_Warfarin ~ Gender + Race + Ethnicity + Age + Height +
               Weight + INR + CYP2C9 + VKORC1 + amiodarone +
               aspirin + statin + azoles + sulfa + acetaminophen +
               clopidogrel + enzyme_inducer, data = mydata)
summary(model1)
exp(coefficients(model1))
```
Adjusted R-squared:  0.4524

```{r} 
model2 <- lm(Dose_Warfarin ~ Gender + Race + Ethnicity + Age + Height + Weight +
               INR + CYP2C9 + VKORC1 + amiodarone +
               azoles + enzyme_inducer, data = mydata)
summary(model2)
exp(coefficients(model2))
```
Adjusted R-squared:  0.453 


#Step function 1
```{r} 
fit_null <- lm(Dose_Warfarin~1, data = mydata)
fit_full <- lm(Dose_Warfarin ~  Gender + Race + Ethnicity + Age + Height + Weight +
               INR + CYP2C9 + VKORC1 + amiodarone +
               azoles + enzyme_inducer, data = mydata)
fit.step <- step(fit_null, scope=list(lower=fit_null, upper=fit_full), direction= "both")
```
Step:  AIC=20193.87
Same as MODEL 2

Replace response variable therapeutic warfarin dose with square root of therapeutic warfarin dose.
```{r} 
mydata$root_warfarin <- (mydata$Dose_Warfarin)^0.5
model3 <- lm(root_warfarin ~ VKORC1 + Age + Weight + CYP2C9 + amiodarone + 
    INR + Ethnicity + Height + azoles + enzyme_inducer + Gender + 
    Race, data = mydata)
summary(model3)
exp(coefficients(model3))
```
Adjusted R-squared:  0.5133


#7. Fit an SVM model to classify patients into categories of high/low warfarin dosage.
```{r} 
#install.packages("e1071", dependencies = TRUE)
library("e1071")

mydata2 <- rbind(HighDose,LowDose)
mydata2$warfarin_level <- as.character(mydata2$warfarin_level)
mydata2$warfarin_level <- as.factor(mydata2$warfarin_level)

mydata2 <- na.omit(mydata2)
attach(mydata2)
x <- subset(mydata2, select = -warfarin_level)
x <- x[,-7]
y <- warfarin_level

#Test your results
n <- nrow(mydata2)
ntrain <- round(n*0.8) # number of training examples
tindex <- sample(n,ntrain) # indices of training samples
trainset <- x[tindex,]
testset <- x[-tindex,]
ytrain <- y[tindex]
ytest <- y[-tindex]

svmfit <- svm(ytrain ~ ., data = trainset, kernel = "radial",cost=10, gamma=0.5, probability = TRUE, scale = TRUE)
summary(svmfit)
```
Plot SVM - Weight ~ Height
```{r} 
x2 <- mydata2[,5:6]
y2 <- mydata2[,19]
trainset2 <- x2[tindex,]
testset2 <- x2[-tindex,]
ytrain2 <- y2[tindex]
ytest2 <- y2[-tindex]

svmfit2 <- svm(ytrain2 ~ ., data = trainset2, kernel = "radial",cost=10, gamma=0.5, probability = TRUE, scale = TRUE)
print(svmfit2)

plot(svmfit2, trainset2, Weight ~ Height)
```
Predict labels on test
```{r} 
ypred = predict(svmfit,testset, probability = TRUE)
system.time(predict(svmfit,testset))
```
The confusion matrix result of prediction
```{r} 
table(ypred,ytest)
```
Compute at the prediction scores
```{r} 
ypredscore = predict(svmfit,testset,type="decision")
```
Compute accuracy
```{r} 
sum(ypred==ypredscore)/length(ypredscore)
```

Check that the predicted labels are the signs of the scores
```{r} 
table(ypredscore,ypred)
```
#ROC curve
```{r} 
#install.packages("ROCR")
#install.packages("pROC")
library(ROCR)
library(pROC)

#Change warfarin level to numeric and predict again
mydata2$warfarin_level <- as.character(mydata2$warfarin_level)
mydata2$warfarin_level[mydata2$warfarin_level == "High dose"] <- 1
mydata2$warfarin_level[mydata2$warfarin_level == "Low dose"] <- 0
mydata2$warfarin_level <- as.numeric(mydata2$warfarin_level)

mydata2 <- na.omit(mydata2)
attach(mydata2)
x <- subset(mydata2, select = -warfarin_level)
x <- x[,-7]
y <- warfarin_level
n <- nrow(mydata2)
ntrain <- round(n*0.8) # number of training examples
tindex <- sample(n,ntrain) # indices of training samples
trainset <- x[tindex,]
testset <- x[-tindex,]
ytrain <- y[tindex]
ytest <- y[-tindex]

svmfit <- svm(ytrain ~ ., data = trainset, kernel = "radial",cost=10, gamma=0.5, probability = TRUE, scale = TRUE)

ypred = predict(svmfit,testset, probability = TRUE)
ypredscore = predict(svmfit,testset,type="decision")

rocplot <- function(pred, truth, ...) {
  predob = prediction(pred, truth)
  perf = performance(predob, "tpr", "fpr")
  plot(perf, ...)
  area <- auc(truth, pred)
  area <- format(round(area, 4), nsmall = 4)
  text(x=0.8, y=0.1, labels = paste("AUC =", area))

  # the reference x=y line
  segments(x0=0, y0=0, x1=1, y1=1, col="gray", lty=2)
}

rocplot(ypredscore, ytest, col="blue")
```
Plot precision/recall curve
```{r} 
pred <- prediction(ypredscore,ytest)
perf <- performance(pred, measure = "prec", x.measure = "rec")
plot(perf)
```
Plot accuracy as function of threshold
```{r} 
perf <- performance(pred, measure = "acc")
plot(perf)
```

#8. Briefly summarize your findings, including a few sentences comparing the summary results and a plot (ex: Residuals vs Fitted and/or Normal Q-Q) between the two regression models. Also briefly discuss the SVM classifier.
```{r} 
par(mfrow=c(1,1))
plot(model2, pch=16, which=1)
```

```{r}
par(mfrow=c(1,1))
plot(model3, pch=16, which=1)
```

##__Summary Report__
#Model 2
Step:  AIC=20193.87 , Adjusted R-squared:  0.453

According to the results from Model 2, linear regression was calculated to predict warfarin dose based on VKORC1, Age, Weight, CYP2C9, amiodarone, INR, Ethnicity, Height, azoles, enzyme_inducer, Gender, Race

A significant regression equation was found : Dose_Warfarin = 2.47 + 0.24(Gendermale) + 0.05(Age) + 1.12(Height) + 1.18(Weight) + 2.31(INR) + 1.48e-08(CYP2C9*1/*11) + 0.01(CYP2C9*1/*2) + 7.88e-05(CYP2C9*1/*3) + 4.69e-07(CYP2C9*1/*5) + 2.95e-06(CYP2C9*2/*2) + 4.08e-09(CYP2C9*2/*3) + 4.86e-10(CYP2C9*3/*3) + 3.16e-05(VKORC1A/G) + 2.60e-09(VKORC1A/A) + 6.17e-04(amiodarone1) + 0.03(azoles1) + 3.7(enzyme_inducer1)

#Model 3
Adjusted R-squared:  0.5133

According to the results from Model 3, linear regression was calculated to predict square root of warfarin dose based on VKORC1, Age, Weight, CYP2C9, amiodarone, INR, Ethnicity, Height, azoles, enzyme_inducer, Gender, Race

A significant regression equation was found : root_warfarin = 43.58 + 0.43(VKORC1A/G) + 0.17(VKORC1A/A) + 0.78(Age) + 1.01(Weight) + 0.22(CYP2C9*1/*11) + 0.60(CYP2C9*1/*2) + 0.42(CYP2C9*1/*3) + 0.34(CYP2C9*1/*5) + 0.31(CYP2C9*2/*2) + 0.15(CYP2C9*2/*3) + 0.10(CYP2C9*3/*3) + 0.52(amiodarone1) +  1.38(INR) + 1.01(Height) + 0.80(azoles1) + 0.86(Gendermale) + 0.79(RaceBlack or African American)


Model 3 has adjusted R-squared 0.5133 that means this model can explain the relationship between predictors and warfarin dose 51.33 %. 


Residuals vs Fitted plot of Model 5: <br />
1. The relationship is linear is reasonable. <br />
2. The residuals roughly form a "horizontal band"  around the 0 line. This suggests that the variances of the error terms are equal. <br />
3. There are some outliers (4247, 4163, 3903). <br />

#SVM
For SVM classification plot, patients who have high height and high weight trend to use high dose of warfarin. <br />

After fitting SVM model with using Radial Kernel, ROC curve is obtained by changing the threshold 0 to threshold t in f(X) > t, and recording false positive and true positive rates as t varies. Area under the curve (AUC) is good.

